#hmwk1

## data input

### load packages
```{r}
library(sf)
library(tmap) 
library(tmaptools)
library(RSQLite)
library(tidyverse)
```

### read in the shape file
```{r}
library(sf)
shape_hmwk1 <- st_read("1_2_hmwk1_data/statsnz-territorial-authority-2018-generalised-SHP/territorial-authority-2018-generalised.shp")
```

### see a summary of the shape data
```{r}
summary(shape_hmwk1)
```

### view the shape file
```{r}
plot(shape_hmwk1)
```

### view the shape geometry
```{r}
library(sf)
shape_hmwk1 %>% 
  st_geometry() %>%
  plot()
```

### simplify geometry
```{r}
shape_simple <- st_simplify(shape_hmwk1, dTolerance = 1000)
```

### view the shape_simple file
```{r}
plot(shape_simple)
```

### view the shape_simple geometry
```{r}
library(sf)
shape_simple %>% 
  st_geometry() %>%
  plot()
```

### read in the csv
```{r}
library(tidyverse)
employed_2018  <-  read_csv("1_2_hmwk1_data/2018_SA1_dataset_TerritorialAuthority_edited.csv")
```

### view the .csv
```{r}
employed_2018 
```

## data manuipulation

### added test to last row to make character for example
```{r}
Datatypelist <- shape_simple %>% 
  summarise_all(class)
Datatypelist
```

### join the data
```{r}
  shape2 <- shape_simple%>%
  merge(.,
        employed_2018,
        by.x="TA2018_V_1", 
        by.y="Area_Description")
```

### option to join on the id by making the column numeric
```{r}
shape3 <- shape_simple %>%
  mutate(TA2018_V1_=(as.numeric(TA2018_V1_)))%>%
  merge(.,
        employed_2018,
        by.x="TA2018_V1_", 
        by.y="Area_Code")
```

### check the join
```{r}
shape3%>%  
head(., n=10)
```

## data mapping

### set tmap to plot  
```{r}
  library(tmap)
tmap_mode("plot")
```

### view the map
```{r}
shape2 %>%
qtm(.,fill = "Paid employee")
```

### write a .gpkg
```{r}
shape %>%
  st_write(.,"1_2_hmwk1_outputs/hmwk1.gpkg","2018_paid_employee",
           delete_layer=TRUE)
```

### connect to the .gpkg
```{r}
library(readr)
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(),dbname="1_2_hmwk1_outputs/hmwk1.gpkg")
```

### list contents of .gpkg
```{r}
con %>%
  dbListTables()
```

### add the original .csv
```{r}
con %>%
  dbWriteTable(.,
               "original_csv",
               employed_2018,
               overwrite=TRUE)
```

### disconnect from .gpkg
```{r}
con %>% 
  dbDisconnect()
```
