#prac2

## task
By the end of this practical you should be able to:

Execute basic processing in R
Examine, clean and manipulate comma seperate value (.csv) data
Examine, clean and manipulate and plot spatial (.shp) data
Produce interactive maps
Evaluate the benefits of different data manipulation and mapping techniques

## data input

### install packages
```{r}
install.packages("here")
install.packages("plotly")
install.packages("RTools")
install.packages("maptools")
install.packages(c("classInt", "tmap"))
install.packages(c("RColorBrewer", "sp", "rgeos", 
                   "tmaptools", "sf", "downloader", "rgdal", 
                   "geojsonio"))
```

### load packages
```{r}
library(sf)
library(tmap) 
library(tmaptools)
library(RSQLite)
library(tidyverse)
library(dplyr)
library(here)
library(janitor)
library(readr)
library(plotly)
library(maptools)
library(RColorBrewer)
library(classInt)
library(sp)
library(rgeos)
library(tmap)
library(tmaptools)
library(sf)
library(rgdal)
library(geojsonio)
```

### read in the csv OSC

```{r}
LondonDataOSC<- read.csv("2_1_prac2_data/ward-profiles-excel-version.csv", 
                         header = FALSE, 
                         sep = ",",  
                         encoding = "latin1")
```

### read in the csv

```{r}
LondonData <- read_csv("https://data.london.gov.uk/download/ward-profiles-and-atlas/772d2d64-e8c6-46cb-86f9-e52b4c7851bc/ward-profiles-excel-version.csv", 
                       locale = locale(encoding = "latin1"))
```

### read in the csv and clean

```{r}
LondonData <- read_csv("https://data.london.gov.uk/download/ward-profiles-and-atlas/772d2d64-e8c6-46cb-86f9-e52b4c7851bc/ward-profiles-excel-version.csv",
                       locale = locale(encoding = "latin1"),
                       na = "n/a")
```
### view the .csv
```{r}
LondonData 
```
### examine the data type

```{r}
class(LondonData)
```
### check the data has been read in correctly

```{r}
Datatypelist <- LondonData %>% 
  summarise_all(class) %>%
  pivot_longer(everything(), 
               names_to="All_variables", 
               values_to="Variable_class")

Datatypelist
```

### view the column headers

```{r}
LondonData%>%
  colnames()%>%
  # just look at the head, top5
  head()
```

## data wrangling

### selecting the data subset needed
```{r}
LondonBoroughs<-LondonData[626:658,]
```

### or slicing the data subset needed
```{r}
LondonBoroughs<-LondonData%>%
  slice(626:658)
```

### filter based on a variable - female life expectancy greater than 90
```{r}
Femalelifeexp<- LondonData %>% 
  filter(`Female life expectancy -2009-13`>90)
```

### or fliter with stringer when data is charachter not numeric
```{r}
LondonBoroughs<- LondonData %>% 
  filter(str_detect(`New code`, "^E09"))
```

### check it worked

```{r}
LondonBoroughs$`Ward name`
```

### make the rows distinct

```{r}
LondonBoroughs<-LondonBoroughs %>%
  distinct()
```

### selecting columns
```{r}
LondonBoroughs_manualcols<-LondonBoroughs[,c(1,19,20,21)]
```

### alternative way using dplyr with select()
```{r}
LondonBoroughs_dplyrcols<-LondonBoroughs %>%
  dplyr::select(c(1,19,20,21))
```

### selecting columns using words
```{r}
LondonBoroughs_dplyrcols<-LondonBoroughs %>%
LondonBoroughs_contains<-LondonBoroughs %>% 
  dplyr::select(contains("expectancy"), 
         contains("obese - 2011/12 to 2013/14"),
         contains("Ward name")) 
```
### tidying up column names
```{r}
LondonBoroughs <- LondonBoroughs %>%
  dplyr::rename(Borough=`Ward name`)%>%
  clean_names()
```

### using mutate to add new variables based on existing ones
```{r}
Life_expectancy <- LondonBoroughs %>% 
  #new column with average of male and female life expectancy
  mutate(averagelifeexpectancy= (female_life_expectancy_2009_13 +
                                    male_life_expectancy_2009_13)/2)%>%
  #new column with normalised life expectancy
  mutate(normalisedlifeepectancy= averagelifeexpectancy /
           mean(averagelifeexpectancy))%>%
  #select only columns we want
  dplyr::select(new_code,
         borough,
         averagelifeexpectancy, 
         normalisedlifeepectancy)%>%
  #arrange in descending order
  #ascending is the default and would be
  #arrange(normalisedlifeepectancy)
  arrange(desc(normalisedlifeepectancy))
```

### use dplyr to show us the top and bottom number of rows
```{r}
#top of data
slice_head(Life_expectancy, n=5)

#bottom of data
slice_tail(Life_expectancy,n=5)
```

### use case_when(), to assign a new string depending on if the value is above or below the national average
```{r}
Life_expectancy2 <- Life_expectancy %>%
  mutate(UKcompare = case_when(averagelifeexpectancy>81.16 ~ "above UK average",
                               TRUE ~ "below UK average"))
Life_expectancy2
```
### identifying the range of life expectancies
```{r}
Life_expectancy2_group <- Life_expectancy2 %>%
  mutate(UKdiff = averagelifeexpectancy-81.16) %>%
  group_by(UKcompare)%>%
  summarise(range=max(UKdiff)-min(UKdiff), count=n(), Average=mean(UKdiff))

Life_expectancy2_group
```

### using across() that applies a kind of transformation across the columns selected
```{r}
Life_expectancy3 <- Life_expectancy %>%
  mutate(UKdiff = averagelifeexpectancy-81.16)%>%
  mutate(across(where(is.numeric), round, 3))%>%
  mutate(across(UKdiff, round, 0))%>%
  mutate(UKcompare = case_when(averagelifeexpectancy >= 81 ~ 
                                 str_c("equal or above UK average by",
                                       UKdiff, 
                                       "years", 
                                       sep=" "), 
                               TRUE ~ str_c("below UK average by",
                                            UKdiff,
                                            "years",
                                            sep=" ")))%>%
  group_by(UKcompare)%>%
  summarise(count=n())
```

### getting the average life expectancy of each London Borough compared to the UK average
```{r}
Life_expectancy4 <- Life_expectancy %>%
  mutate(UKdiff = averagelifeexpectancy-81.16)%>%
  mutate(across(is.numeric, round, 3))%>%
  mutate(across(UKdiff, round, 0))
```
## plotting

### a simple quick plot
```{r}
plot(LondonBoroughs$male_life_expectancy_2009_13,
     LondonBoroughs$percent_children_in_reception_year_who_are_obese_2011_12_to_2013_14)
```

### using ploty to enhance the graphics of the graph
```{r}
plot_ly(LondonBoroughs, 
        #data for x axis
        x = ~male_life_expectancy_2009_13, 
        #data for y axis
        y = ~percent_children_in_reception_year_who_are_obese_2011_12_to_2013_14, 
        #attribute to display when hovering 
        text = ~borough, 
        type = "scatter", 
        mode = "markers")
```

## spatial data input

### read in the shape file
```{r}
EW <- st_read(here::here("2_1_prac2_data","LAD_Dec_2015_GCB_GB.shp"))
```

### see a summary of the shape data
```{r}
summary(EW)
```

### view the shape file
```{r}
plot(EW)
```

### pull out the london boundaries using str_detect in combination with filter
```{r}
LondonMap<- EW %>%
  filter(str_detect(lad15cd, "^E09"))
```


### plot the london map
```{r}
qtm(LondonMap)
```

## joining

### join the data
```{r}
  BoroughDataMap2 <- EW %>% 
  clean_names() %>%
  filter(str_detect(lad15cd, "^E09"))%>%
  left_join(., 
            LondonData,
            by = c("lad15cd" = "new_code"))
```

### make the rows distinct
```{r}
BoroughDataMap2<-BoroughDataMap2 %>%
  distinct()
```

## simple mapping

### make a qtm from tmap  
```{r}
library(tmap)
library(tmaptools)
tmap_mode("plot")
qtm(BoroughDataMap2, 
    fill = "rate_of_job_seekers_allowance_jsa_claimants_2015")
```

### adding a basemap
```{r}
tmaplondon <- BoroughDataMap2 %>%
  st_bbox(.) %>% 
  tmaptools::read_osm(., type = "osm", zoom = NULL)
```

### make a map from tmap
```{r}
tmap_mode("plot")

tm_shape(tmaplondon)+
tm_rgb()+
tm_shape(BoroughDataMap2) + 
tm_polygons("rate_of_job_seekers_allowance_jsa_claimants_2015", 
        style="pretty",
        palette="YlOrBr",
        midpoint=NA,
        title="Rate per 1,000 people",
        alpha = 0.5) + 
  tm_compass(position = c("left", "bottom"),type = "arrow") + 
  tm_scale_bar(position = c("left", "bottom")) +
  tm_layout(title = "Job seekers' Allowance Claimants", legend.position = c("right", "bottom"))
```
