#hmwk2

## task
The task is to join some non spatial data to some spatial data and wrangle it.

You need calcuate the average percent of science students (in all) grades per county meeting the required standards and produce a map to show where the Country averages are above or below the State of Washington average.

Download Washing Counties

Download school report card data

Tip do not try and pull out the number from the % symbol, you have the count of students who expected to test and the count who met standard, so you can calculate the percentage.

## data input

### install packages
```{r}

```

### load packages
```{r}
library(sf)
library(tidyverse)
library(here)
library(janitor)
library(tmap) 
```

### read in the csv and remove null as no data
```{r}
report <- read_csv(here::here("2_2_hmwk2_data", 
                   "Report_Card_Assessment_Data_2018-19_School_Year_20231030.csv"),
                              na= "NULL")
```

### view the .csv
```{r}
report 
```

### read in the shape file
```{r}
shape_hmwk2 <- st_read(here::here("2_2_hmwk2_data",                          "Washington_Counties_with_Natural_Shoreline___washsh_area",                           "Washington_Counties_with_Natural_Shoreline___washsh_area.shp"))
```

## data wrangling - calculate the average percent of science students (in all) grades per county meeting the required standards 1 count met standard, 2 count expected to test, 3 test subject

### wrangle 1 - summarise class and pivot longer 
```{r}
Datatypelist <- report %>% 
  summarise_all(class) %>%
  pivot_longer(everything(), 
               names_to="All_variables", 
               values_to="Variable_class")
```

### wrangle 1 - select columns: country, organisation_level subject, grade, count the met standard, count those tested. then filter for county to omit multiple, organisation level as school, test subject as science and all grade levels. clean up to remove n/as. then mutate to work out % that have met 
```{r}
county_only <- report %>%
  clean_names(.) %>%
  select(county, organization_level, test_subject, count_met_standard, 
         count_of_students_expected_to_test, grade_level)%>%
  # the != means don't select this, but select everything else
  # i could also filter on where 
  filter(county != "Multiple")%>%
  filter(organization_level == "School")%>%
  filter(test_subject == "Science")%>%
  filter(grade_level=="All Grades")%>%
    group_by(county)%>%
  # we need to remove NAs - note we can use this function or do it within the summarise below with the argument na.rm=T they do the same thing!
  na.omit()%>%
  # na.rm = T means remove missing values from the data
  # could also use na.omit or filter greater than 0
  summarise(total_county_met_standard=sum(count_met_standard), 
            total_county_to_test=sum(count_of_students_expected_to_test))%>%
  mutate(percent_met_per_county=(total_county_met_standard/total_county_to_test)*100)
```

### wrangle 2 - to work out what the state average is by averaging the percent_met_per_county column from the county_only object
```{r}
state_average <- county_only%>%
  mutate(sum_met=sum(total_county_met_standard))%>%
  mutate(sum_to_test =sum(total_county_to_test))%>%
  mutate(state_average=((sum_met/sum_to_test)*100))%>%
  head(n=1)%>%
  pull(state_average)
```

### wrangle 3 - compare the county average to the state average by subtracting the state average from each county average to work out which counties are above or below than the state average. make a new column that records which county is above or below and assign through case_when to identify this county by county
```{r}
county_only_above_below_state <- county_only %>%
  mutate(difference_to_state=(percent_met_per_county-state_average))%>%
  mutate(across(difference_to_state , round, 0))%>%
  mutate(above_below = case_when(difference_to_state<0 ~ "below",
                                   difference_to_state>0 ~ "above",
                                   difference_to_state==0 ~ "equal"
                                   ))
```

## join the data

### make a new object and left join the latest non-spatial data object made with the spatial data
```{r}
joined_data_hmwk2 <- shape_hmwk2 %>% 
  clean_names(.) %>%
  left_join(., 
            county_only_above_below_state,
            by = c("countylabe" = "county"))
```

### view the new joined data
```{r}
joined_data_hmwk2
```

## data mapping

### set tmap to plot  
```{r}
tmap_mode("plot")
```

### from the grammer of graphics, specify your spatial data and then add aesthetics
```{r}
#bbox_county <- joined_data_hmwk2 %>%
#  st_bbox(.) %>% 
#  tmaptools::read_osm(., type = "osm", zoom = NULL)

#tm_shape(bbox_county)+
#  tm_rgb()+

  tm_shape(joined_data_hmwk2) + 
  tm_polygons("above_below", 
             # style="pretty",
              palette="Blues",
              midpoint=NA,
              #title="Number of years",
              alpha = 0.5) + 
  tm_compass(position = c("left", "bottom"),type = "arrow") + 
  tm_scale_bar(position = c("left", "bottom")) +
  tm_layout(title = "Counties above or below state avearge for science in all grades", 
            legend.position = c("right", "bottom"))
```

