#prac3

## task
By the end of this practical you should be able to:

Load, manipulate and interpret raster layers
Observe and critique different descriptive data manipulation methods and outputs

## data input

### install packages
```{r}
install.packages(c("fs", "terra", "raster", "ggplot"))
```

### load packages
```{r}
library(sf)
library(here)
library(janitor)
library(tidyverse)
library(RSQLite)
library(fs)
library(raster)
library(terra)
library(ggplot2)
```
### read in the vector data for Australia
```{r}
st_layers(here("3_1_prac3_data", "gadm36_AUS_gpkg", "gadm36_AUS.gpkg"))
```
### read in the GeoPackage layer for the whole of Australia (layer ending in 0)
```{r}
Ausoutline <- st_read(here("3_1_prac3_data", "gadm36_AUS_gpkg", "gadm36_AUS.gpkg"), 
                      layer='gadm36_AUS_0')
```

### check the coordinate reference system
```{r}
print(Ausoutline)
```

### using proj4 string to identify the CRS
```{r}
st_crs(Ausoutline)$proj4string
```

### reprojecting the australian outline from WGS84 to GDA94, which is a local CRS for Australia
```{r}
AusoutlinePROJECTED <- Ausoutline %>%
  st_transform(.,3112)

print(AusoutlinePROJECTED)
```

### read in part of the raster data
```{r}
jan<-terra::rast(here("3_1_prac3_data", "wc2.1_5m_tmax", "wc2.1_5m_tmax_01.tif"))
```

### view that part of the raster data
```{r}
jan
```

### plot that part of the raster data
```{r}
plot(jan)
```

### search our directory for all file names containing.tif
```{r}
library(fs)
dir_info("3_1_prac3_data/wc2.1_5m_tmax") 
```

### list the files you want
```{r}
listfiles<-dir_info("3_1_prac3_data/wc2.1_5m_tmax") %>%
  filter(str_detect(path, ".tif")) %>%
  dplyr::select(path)%>%
  pull()
```

### have a look at the selected list
```{r}
listfiles
```

### load all of the data into a SpatRaster which is a collection of raster layers
```{r}
worldclimtemp <- listfiles %>%
  terra::rast()
```

### view the SpatRaster
```{r}
worldclimtemp
```

### to access just one of the layers
```{r}
worldclimtemp[[1]]
```

### rename the layers within the stack
```{r}
month <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
           "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

names(worldclimtemp) <- month
```

### now access just one of the layers using the renamed layers
```{r}
worldclimtemp$Jan
```
## extracting data

### make a dataframe of some sample sites — Australian cities/towns
```{r}
site <- c("Brisbane", "Melbourne", "Perth", "Sydney", "Broome", "Darwin", "Orange", 
          "Bunbury", "Cairns", "Adelaide", "Gold Coast", "Canberra", "Newcastle", 
          "Wollongong", "Logan City" )
lon <- c(153.03, 144.96, 115.86, 151.21, 122.23, 130.84, 149.10, 115.64, 145.77, 
         138.6, 153.43, 149.13, 151.78, 150.89, 153.12)
lat <- c(-27.47, -37.91, -31.95, -33.87, 17.96, -12.46, -33.28, -33.33, -16.92, 
         -34.93, -28, -35.28, -32.93, -34.42, -27.64)
#Put all of this inforamtion into one list 
samples <- data.frame(site, lon, lat, row.names="site")
# Extract the data from the Rasterstack for all points 
AUcitytemp<- terra::extract(worldclimtemp, samples)
```

### add the city names to the rows of AUcitytemp
```{r}
Aucitytemp2 <- AUcitytemp %>% 
  as_tibble()%>% 
  add_column(Site = site, .before = "Jan")
```

## make some descriptive statistics

### data preparation
```{r}
Perthtemp <- Aucitytemp2 %>%
  filter(site=="Perth")
```

### make a histogram
```{r}
hist(as.numeric(Perthtemp))
```

### add aesthetics to the histogram
```{r}
#define where you want the breaks in the historgram
userbreak<-c(8,10,12,14,16,18,20,22,24,26)

# remove the ID and site columns
Perthtemp <- Aucitytemp2 %>%
  filter(site=="Perth")

t<-Perthtemp %>%
 dplyr::select(Jan:Dec)

  hist((as.numeric(t)), 
     breaks=userbreak, 
     col="red", 
     main="Histogram of Perth Temperature", 
     xlab="Temperature", 
     ylab="Frequency")
```

### view the histogram info that R generated
```{r}
histinfo <- as.numeric(t) %>%
  as.numeric()%>%
  hist(.)
```
### view the info
```{r}
histinfo
```

## making a more complex histogram based on the distribution of temperatures for the whole of Australia in Jan

### plot the geometry of Australia
```{r}
plot(Ausoutline$geom)
```

### simplify the geometry
```{r}
AusoutSIMPLE <- Ausoutline %>%
  st_simplify(., dTolerance = 1000) %>%
  st_geometry()%>%
  plot()
```
### check the CRS for the outline of Australia
```{r}
print(Ausoutline)
```

### check the CRS for the raster temp data
```{r}
crs(worldclimtemp)
```

### set the map extent of the temp data based on Ausoutline
```{r}
Austemp <- Ausoutline %>%
  # now crop our temp data to the extent
  terra::crop(worldclimtemp,.)

# plot the output
plot(Austemp)
```

###  use the mask()function to cookie cutter the raster with the vector layer
```{r}
exactAus<-terra::mask(Austemp, Ausoutline)
```

### recompute the histogram for australia in march
```{r}
hist(exactAus[[3]], col="red", main ="March temperature")
```

### converting raster to a dataframe for making a histogram with ggplot
```{r}
exactAusdf <- exactAus %>%
  as.data.frame()
```

### using ggplot for making a histogram with more control
```{r}
# set up the basic histogram
gghist <- ggplot(exactAusdf, 
                 aes(x=Mar)) + 
  geom_histogram(color="black", 
                 fill="white")+
  labs(title="Ggplot2 histogram of Australian March temperatures", 
       x="Temperature", 
       y="Frequency")
# add a vertical line to the hisogram showing mean tempearture
gghist + geom_vline(aes(xintercept=mean(Mar, 
                                        na.rm=TRUE)),
            color="blue", 
            linetype="dashed", 
            size=1)+
  theme(plot.title = element_text(hjust = 0.5))
```

## making a histogram for multiple months

### turn our variable (months) into long data
```{r}
squishdata<-exactAusdf%>%
  pivot_longer(
  cols = 1:12,
  names_to = "Month",
  values_to = "Temp"
)
```

### subset the data, selecting two months
```{r}
twomonths <- squishdata %>%
  # | = OR
  filter(., Month=="Jan" | Month=="Jun")
```

### get the mean for each month we selected
```{r}
meantwomonths <- twomonths %>%
  group_by(Month) %>%
  summarise(mean=mean(Temp, na.rm=TRUE))

meantwomonths
```

### select the colour and fill based on the variable
```{r}
ggplot(twomonths, aes(x=Temp, color=Month, fill=Month)) +
  geom_histogram(position="identity", alpha=0.5)+
  geom_vline(data=meantwomonths, 
             aes(xintercept=mean, 
                 color=Month),
             linetype="dashed")+
  labs(title="Ggplot2 histogram of Australian Jan and Jun
       temperatures",
       x="Temperature",
       y="Frequency")+
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5))
```

### fix error message by dropping all the NAs and make sure month columns have levels specified so it will map in decending order also select a bin width of 5
```{r}
data_complete_cases <- squishdata %>%
  drop_na()%>% 
  mutate(Month = factor(Month, levels = c("Jan","Feb","Mar",
                                          "Apr","May","Jun",
                                          "Jul","Aug","Sep",
                                          "Oct","Nov","Dec")))

# Plot faceted histogram
ggplot(data_complete_cases, aes(x=Temp, na.rm=TRUE))+
  geom_histogram(color="black", binwidth = 5)+
  labs(title="Ggplot2 faceted histogram of Australian temperatures", 
       x="Temperature",
       y="Frequency")+
  facet_grid(Month ~ .)+
  theme(plot.title = element_text(hjust = 0.5))
```
### making an interactive histogram using plotly
```{r}
library(plotly)
# split the data for plotly based on month

jan <- squishdata %>%
  drop_na() %>%
  filter(., Month=="Jan")

jun <- squishdata %>%
  drop_na() %>%
  filter(., Month=="Jun")

# give axis titles
x <- list (title = "Temperature")
y <- list (title = "Frequency")

# set the bin width
xbinsno<-list(start=0, end=40, size = 2.5)

# plot the histogram calling all the variables we just set
ihist<-plot_ly(alpha = 0.6) %>%
        add_histogram(x = jan$Temp,
        xbins=xbinsno, name="January") %>%
        add_histogram(x = jun$Temp,
        xbins=xbinsno, name="June") %>% 
        layout(barmode = "overlay", xaxis=x, yaxis=y)

ihist
```
## other descriptive statistics to consider

### mean per month
```{r}
meanofall <- squishdata %>%
  group_by(Month) %>%
  summarise(mean = mean(Temp, na.rm=TRUE))

# print the top 1
head(meanofall, n=1)
```
### standard deviation per month
```{r}
sdofall <- squishdata %>%
  group_by(Month) %>%
  summarize(sd = sd(Temp, na.rm=TRUE))
```

### maximum per month
```{r}
maxofall <- squishdata %>%
  group_by(Month) %>%
  summarize(max = max(Temp, na.rm=TRUE))
```

### minimum per month
```{r}
minofall <- squishdata %>%
  group_by(Month) %>%
  summarize(min = min(Temp, na.rm=TRUE))
```

### interquartlie range per month
```{r}
IQRofall <- squishdata %>%
  group_by(Month) %>%
  summarize(IQR = IQR(Temp, na.rm=TRUE))
```

### perhaps you want to store multiple outputs in one list
```{r}
lotsofstats <- squishdata %>%
  group_by(Month) %>%
  summarize(IQR = IQR(Temp, na.rm=TRUE), 
            max=max(Temp, na.rm=T))
```

### or you want to know the mean (or some other stat) for the whole year as opposed to each month
```{r}
meanwholeyear=squishdata %>%
  summarize(meanyear = mean(Temp, na.rm=TRUE))
```

