#prac5

## task
By the end of this practical you should be able to:

List and explain basic mapping concepts across QGIS and R
Interpret and manipulate data from multiple sources
Create near publishable static and interactive mapped outputs
Evaluate and critique mapping approaches between QGIS and R

## data input

### install packages
```{r}
install.packages("leafpop")
install.packages("leaflet")
```


### load packages
```{r}
library(sf)
library(tmap) 
library(tmaptools)
library(tidyverse)
library(here)
library(grid)
library(leafpop)
library(leaflet)
```

### read in the london OSM shape file and transform the CRS to BNG
```{r}
OSM <- st_read(here::here("5_1_prac5_data",
                          "greater_london_latest_free_shp", 
                          "gis_osm_pois_a_free_1.shp")) %>%
  st_transform(., 27700) %>%

  filter(fclass == 'hotel')
```

### read in the world cities and transform the CRS to BNG
```{r}
Worldcities <- st_read(here::here("5_1_prac5_data", 
                                  "world_cities", 
                                  "World_Cities.shp")) %>%
  st_transform(., 27700)
```

### read in the uk outline and transform the CRS to BNG
```{r}
UK_outline <- st_read(here::here("5_1_prac5_data", 
                                 "gadm41_GBR_shp", 
                                 "gadm41_GBR_0.shp")) %>%
  st_transform(., 27700)
```

### read in the london boroughs shape file and transform the CRS to BNG 
```{r}
Londonborough <- st_read(here::here("5_1_prac5_data",
                                    "statistical-gis-boundaries-london", 
                                    "ESRI", 
                                    "London_Borough_Excluding_MHW.shp"))%>%
  st_transform(., 27700)
```

### read in the airbnb csv and convert to a simple features object and set x,y and CRS to 4326 and then to BNG, then add in a comparison for the airbnb data and hotel data
```{r}
Airbnb <- read_csv("5_1_prac5_data/listings.csv") %>%
  # longitude is considered x value here, latitude is y
  st_as_sf(., coords = c("longitude", "latitude"), 
                   crs = 4326) %>%
    st_transform(., 27700)%>%
    #select entire places that are available all year
    filter(room_type == 'Entire home/apt' & availability_365 =='365')
```
## make functions

### make a join function
```{r}
Joinfun <- function(data1, data2){

output<- data1%>%
  st_join(data2,.) %>%
  add_count(GSS_CODE, name="accomm_in_borough") 

  return(output)}
```

## join the spatial data

### using the join function, make a spatial join of the hotel data to the london boroughs data
```{r}
Hotels <- Joinfun(OSM, Londonborough)
```

### using the join function, make a spatial join of the airbnb data to the london boroughs data
```{r}
Airbnb <- Joinfun(Airbnb, Londonborough)
```

### filter the world cities for UK cities
```{r}
Worldcities2 <- Worldcities %>%
  filter(CNTRY_NAME=='United Kingdom'&
           Worldcities$CITY_NAME=='Birmingham'|
           Worldcities$CITY_NAME=='London'|
           Worldcities$CITY_NAME=='Edinburgh')

newbb <- c(xmin=-296000, ymin=5408, xmax=655696, ymax=1000000)
```

### crop UK outline
```{r}
UK_outlinecrop <- UK_outline$geometry %>%
  st_crop(., newbb)
```

### group_by summaries the count of points for the number of hotels for each london borough
```{r}
Hotels <- Hotels %>%
  group_by(., GSS_CODE, NAME)%>%
  summarise(`Accomodation count` = unique(accomm_in_borough))
```

### group_by summaries the count of points for the number of airbnbs for each london borough
```{r}
Airbnb <- Airbnb %>%
  group_by(., GSS_CODE, NAME)%>%
  summarise(`Accomodation count` = unique(accomm_in_borough))
```

## make a map

### set tmap to plot
```{r}
tmap_mode("plot")
```

### set the breaks for the mapped data
```{r}
breaks = c(0, 5, 12, 26, 57, 381) 
```

### plot each map
```{r}
tm1 <- tm_shape(Hotels) + 
  tm_polygons("Accomodation count", 
              breaks=breaks,
              palette="PuBu")+
  tm_legend(show=FALSE)+
  tm_layout(frame=FALSE)+
  tm_credits("(a)", position=c(0,0.85), size=1.5)

tm2 <- tm_shape(Airbnb) + 
  tm_polygons("Accomodation count",
              breaks=breaks, 
              palette="PuBu") + 
  tm_legend(show=FALSE)+
  tm_layout(frame=FALSE)+
  tm_credits("(b)", position=c(0,0.85), size=1.5)

tm3 <- tm_shape(UK_outlinecrop)+ 
  tm_polygons(col="darkslategray1")+
  tm_layout(frame=FALSE)+
  tm_shape(Worldcities2) +
  tm_symbols(col = "red", scale = .5)+
  tm_text("CITY_NAME", xmod=-1, ymod=-0.5)

legend <- tm_shape(Hotels) +
    tm_polygons("Accomodation count",
                breaks=breaks,
                palette="PuBu") +
    tm_scale_bar(position=c(0.2,0.04), text.size=0.6)+
    tm_compass(north=0, position=c(0.65,0.6))+
    tm_layout(legend.only = TRUE, legend.position=c(0.2,0.25),asp=0.1)+
    tm_credits("(c) OpenStreetMap contrbutors and Air b n b", position=c(0.0,0.0))
  
t=tmap_arrange(tm1, tm2, tm3, legend, ncol=2)

t
```

## map arrangement using the grid package

### move to a new page
```{r}
grid.newpage()
```
### set the viewports
```{r}
pushViewport(viewport(layout=grid.layout(2,2)))
print(tm1, vp=viewport(layout.pos.col=1, layout.pos.row=1, height=5))
print(tm2, vp=viewport(layout.pos.col=2, layout.pos.row=1, height=5))
print(tm3, vp=viewport(layout.pos.col=1, layout.pos.row=2, height=5))
print(legend, vp=viewport(layout.pos.col=2, layout.pos.row=2, height=5))
```

## make an inset map

###  convert it into a class of sf otherwise our bb won't have a class it will just be x and y coordinates for the box this makes it into a polygon
```{r}
Londonbb <- st_bbox(Airbnb,
                    crs = st_crs(Airbnb))%>%
                    st_as_sfc()
```

### set the main map
```{r}
main <- tm_shape(Airbnb, bbbox = Londonbb) + 
  tm_polygons("Accomodation count",
              breaks=breaks, 
              palette="PuBu")+
  tm_scale_bar(position = c("left", "bottom"), text.size = .75)+
  tm_layout(legend.position = c("right","top"), 
            legend.text.size=.75, 
            legend.title.size = 1.1,
            frame=FALSE)+
  tm_credits("(c) OpenStreetMap contrbutors and Air b n b", position=c(0.0,0.0))+
  #tm_text(text = "NAME", size = .5, along.lines =T, remove.overlap=T,  auto.placement=F)+
  tm_compass(type = "8star", position = c(0.06, 0.1)) +

  #bottom left top right
  tm_layout(inner.margin=c(0.02,0.02,0.02,0.2))
```

### set the inset map
```{r}
inset = tm_shape(UK_outlinecrop) + tm_polygons() +
  tm_shape(Londonbb)+ 
  tm_borders(col = "grey40", lwd = 3)+
    tm_layout(frame=FALSE,
            bg.color = "transparent")+
  tm_shape(Worldcities2) +
  tm_symbols(col = "red", scale = .5)+
  tm_text("CITY_NAME", xmod=-1.5, ymod=-0.5)
```

### print the map
```{r}
main
print(inset, vp = viewport(0.86, 0.29, width = 0.5, height = 0.55))
```
## export

### output the map as a png
```{r}
tmap_save(t, 'hotelsandairbnbR.png')

library(grid)
tmap_save(main,insets_tm = inset,insets_vp=viewport(x=0.86, y=0.29, width=.5, height=.55), filename="test.pdf", dpi=600)
```

### make a basic interactive map
```{r}
tmap_mode("view")

tm_shape(Airbnb) + 
  tm_polygons("Accomodation count", breaks=breaks) 
```
## making advanced interactive maps

### join data
```{r}
Joined <- Airbnb%>%
  st_join(., Hotels, join = st_equals)%>%
  dplyr::select(GSS_CODE.x, NAME.x, `Accomodation count.x`, `Accomodation count.y`)%>%
  dplyr::rename(`GSS code` =`GSS_CODE.x`,
                `Borough` = `NAME.x`,
                `Airbnb count` = `Accomodation count.x`,
                `Hotel count`= `Accomodation count.y`)%>%
  st_transform(., 4326)
```

### remove the geometry for our pop up boxes to avoid
```{r}
popupairbnb <-Joined %>%
  st_drop_geometry()%>%
  dplyr::select(`Airbnb count`, Borough)%>%
  popupTable()

popuphotel <-Joined %>%
  st_drop_geometry()%>%
  dplyr::select(`Hotel count`, Borough)%>%
  popupTable()
```

### set tmap to view
```{r}
tmap_mode("view")
```

### set the colour palettes using our previously defined breaks
```{r}
pal1 <- Joined %>%
  colorBin(palette = "YlOrRd", domain=.$`Airbnb count`, bins=breaks)

pal1 <-colorBin(palette = "YlOrRd", domain=Joined$`Airbnb count`, bins=breaks)

pal2 <- Joined %>%
  colorBin(palette = "YlOrRd", domain=.$`Hotel count`, bins=breaks)
```

### make map and add aesthetics
```{r}
map<- leaflet(Joined) %>%
  # add basemap options
  addTiles(group = "OSM (default)") %>%
  addProviderTiles(providers$Stamen.Toner, group = "Toner") %>%
  addProviderTiles(providers$Stamen.TonerLite, group = "Toner Lite") %>%
  addProviderTiles(providers$CartoDB.Positron, group = "CartoDB")%>%
  
  #add our polygons, linking to the tables we just made
  addPolygons(color="white", 
              weight = 2,
              opacity = 1,
              dashArray = "3",
              popup = popupairbnb,
              fillOpacity = 0.7,
              fillColor = ~pal2(`Airbnb count`),
              group = "Airbnb")%>%
  
  addPolygons(fillColor = ~pal2(`Hotel count`), 
              weight = 2,
              opacity = 1,
              color = "white",
              dashArray = "3",
              popup = popupairbnb,
              fillOpacity = 0.7,group = "Hotels")%>%
  # add a legend
  addLegend(pal = pal2, values = ~`Hotel count`, group = c("Airbnb","Hotel"), 
            position ="bottomleft", title = "Accomodation count") %>%
  # specify layers control
  addLayersControl(
    baseGroups = c("OSM (default)", "Toner", "Toner Lite", "CartoDB"),
    overlayGroups = c("Airbnb", "Hotels"),
    options = layersControlOptions(collapsed = FALSE)
  )

map
```

